---
title: "INST_462_Group_Project"
output: html_document
date: "2025-04-15"
---

## Team Members:
Gabriel Gonzalez, Jaida Ries, Joseph Toto, DeJon Young
 

## Roles:
- Gabriel Gonzalez: Developer 
- Jaida Ries: Communicator, Project Manager
- Joseph Toto: Designer
- DeJon Young: Data Engineer

---

## Research Questions:


Our refined research question is:

"How does income correlate with crime rates in Washington, DC, by ward?"

We chose this question because Washington, DC is a city known for disparities between neighborhoods. By focusing on income and crime at the ward level, we aim to investigate whether and how economic inequality may be related to safety and well-being.

Audience: This analysis is especially relevant for policymakers, public health officials, and local residents who are interested in understanding and addressing disparities in community outcomes across DC.

---
## Data Description:

Economic Data:
https://opendata.dc.gov/apps/2f6f21d035984e9c9438ae494022aa2a/explore 

Crime Data:
https://opendata.dc.gov/datasets/dc3289eab3d2400ea49c154863312434_8/explore?location=38.904042%2C-77.012050%2C10.64 

GeoJSON Data of DC Wards: 
https://opendata.dc.gov/datasets/wards-from-2022/explore


The initial datasets selected for Phase 2 are:

1. Economic Characteristics of DC Wards.csv
   - Contains ward-level economic indicators including median household income, poverty rates, and unemployment levels.
   - Key variables:
     - `Ward`: Unique identifier for each ward.
     - `MedianIncome`: Median household income.
     - `PovertyRate`: Percentage of residents below the poverty line.
     - `UnemploymentRate`: Percentage of the population unemployed.

2. Crime_Incidents_in_the_Last_30_Days.csv 
   - Includes crime incidents reported in the last 30 days, categorized by ward.
   - Key variables:
     - `WARD`: Unique identifier for each ward.
     - `OFFENSE`: Type of crime committed.
     - `REPORT_DAT`: Date and time when the crime was reported.

---

Before we can start visualizing the data, we need to clean it up a bit.
## Code 

```{r}
library(tidyverse)
library(ggplot2)
library(sf)
library(tmap)
library(RColorBrewer)
```

# Loading and Cleaning Economic Data
```{r}
# Load Economic Data
economic_data <- read_csv("Economic Characteristics of DC Wards.csv", show_col_types = FALSE ) %>%
  select(
    Ward = Name,
    MedianIncome = `INCOME AND BENEFITS (IN 2022 INFLATION-ADJUSTED DOLLARS):  Total households:  Median household income (dollars)`,
    PovertyRate = `PERCENTAGE WHOSE INCOME IN THE PAST 12 MONTHS IS BELOW THE POVERTY LEVEL:  All people`,
    UnemploymentRate = `EMPLOYMENT STATUS:  Civilian labor force:  Unemployment Rate`
  ) %>%
  mutate(Ward = str_extract(Ward, "Ward \\d"))  # Ensure consistent formatting

# Display the first few rows
head(economic_data)
```

# Loading and Aggregating Crime Data
```{r}
crime_data <- read_csv("Crime_Incidents_in_the_Last_30_Days.csv", show_col_types = FALSE)

# Aggregate Crime Data by Ward 
aggregated_crime <- crime_data %>%
  group_by(WARD) %>%
  summarize(TotalCrimes = n()) %>%
  ungroup()

# Display the aggregated data
head(aggregated_crime)

```

# Code to Modify the Ward Column 
```{r}
economic_data_cleaned <- economic_data %>%
  mutate(Ward = str_extract(Ward, "\\d+"))

# View the cleaned data 
print(economic_data_cleaned)

```

# Merging the Datasets 
```{r}

# Convert Ward in economic_data_cleaned to numeric 
economic_data_cleaned <- economic_data_cleaned %>%
  mutate(Ward = as.numeric(Ward))

# Merging the Datasets
merged_data <- economic_data_cleaned %>%
  inner_join(aggregated_crime, by = c("Ward" = "WARD"))

# View the merged data 
print(merged_data)
```


Now that our data is ready to be used, we can now visualize Median Income by Ward, Crimes by Ward, and the relationship between the two. 

First, we created this map of DC to familiarize ourselves with the locations of each ward.

```{r}
# Load spatial data (GeoJSON)
ward_map_data <- st_read("Wards_from_2022.geojson")

# Visualize the GeoJSON geometries (no choropleth applied yet)
tm_shape(ward_map_data) +
  tm_borders() +
  tm_text ("WARD", size = 0.7, col="black") +
  tm_layout(main.title = "DC Wards Geometries")+
  tm_credits(text = "Source: Open Data DC - Crime Incidents Last 30 Days By Ward - https://opendata.dc.gov/datasets/dc3289eab3d2400ea49c154863312434_8/explore?location=38.904042%2C-77.012050%2C10.64",
             position = c("left", "BOTTOM")
             )

```

Here are some maps to show each ward's median income and crimes (between March and April 2025):

```{r}

# Load and prepare the spatial data
ward_map <- st_read("Wards_from_2022.geojson") %>%
  mutate(WARD = as.numeric(str_remove(NAME, "Ward")))

# Merge with attribute data
ward_map_data <- ward_map %>%
  left_join(merged_data, by = c("WARD" = "Ward"))

# Create the choropleth map
tm_shape(ward_map_data) +
  tm_polygons(
    col = "MedianIncome",
    title = "Median Household Income",
    palette = "magma",
    style = "pretty",
    n = 5,
    legend.hist = TRUE,
    legend.format = list(big.mark = ","),
    border.col = NA  
  ) +
  tm_text("WARD", size = 0.7, col = "white", fontface = "bold") + 
tm_layout(
  title = "Income Distribution Across D.C.: \nDiverse Household Incomes by Area (2022)",
  title.position = c("center", "top"),
  title.size = 1.5,
  frame = FALSE,
  legend.frame = FALSE,
  legend.title.size = 1.0,
  legend.text.size = 0.8,
  
  # Adjust spacing between map and elements inside the plotting area
  inner.margins = c(0.2, 0.02, 0.2, 0.02)  
) +
  tm_credits(
    text = "Source: Open Data DC - Crime Incidents Last 30 Days & Economic Characteristics",
    position = c("left", "bottom"),
    size = 2
  )

```

Here is the same map with crime data

```{r}
tm_shape(ward_map_data) + 
  tm_polygons(
    col = "TotalCrimes",
    title = "Total Crimes", 
    palette = "Reds",         
    style = "pretty",  
    n = 5, 
    legend.hist = TRUE,  
    legend.format = list(big.mark = ",")
  ) + 
  tm_text(
    text = "WARD", 
    size = 0.7, 
    col = "white", 
    fontface = "bold", 
    bg.color = "darkred", 
    bg.alpha = 0.5
  ) +
  tm_layout(
    title = "Crime Distribution Across D.C. Wards: \nConsiderable Differences in Total Incidents (2025)", 
    title.position = c("center", "top"),
    title.size = 1.5,
    frame = FALSE,
    legend.frame = FALSE,
    legend.title.size = 1.0,
    legend.text.size = 0.8,
    
    # Add spacing above and below the map
    inner.margins = c(0.2, 0.02, 0.2, 0.02)
  ) +
  tm_credits(
    text = "Source: Open Data DC - Crime Incidents Last 30 Days",
    position = c("left", "bottom"),
    just = "left",
    size = 0.7
  )

```


Now that we have visualized each ward's crime rates and income levels, we can explore the relationship between the two variables.


Median Household Income vs. Total Crimes Plot
```{r}

ggplot(merged_data, aes(x = MedianIncome, y = TotalCrimes)) + 
  geom_point(color = brewer.pal(8, "Set2")[1], size = 3) +  
  geom_smooth(method = "lm", se = FALSE, color = brewer.pal(9, "Blues")[6]) +  
  geom_text(aes(label = Ward), vjust = -1, size = 3, color = "black") +  
  labs(title = "Median Household Income vs. Total Crimes by Ward",
       x = "Median Household Income ($)",
       y = "Total Crimes",
       caption = "Source: Open Data DC - Crime Incidents Last 30 Days") + 
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.caption = element_text(hjust = 0, size = 10)
  )
```



Poverty Rate vs. Total Crimes Plot
```{r}
ggplot(merged_data, aes(x = PovertyRate, y = TotalCrimes)) +
  geom_point(color = brewer.pal(8, "Set2")[2], size = 3) +  
  geom_smooth(method = "lm", se = FALSE, color = brewer.pal(9, "Blues")[5]) +  
  geom_text(aes(label = Ward), vjust = -1, size = 3, color = "black") +  
  labs(title = "Poverty Rate vs. Total Crimes by Ward",
       x = "Poverty Rate (%)",
       y = "Total Crimes",
       caption = "Source: Open Data DC - Crime Incidents Last 30 Days") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.caption = element_text(hjust = 0, size = 10)
  )

```


Unemployment Rate vs. Total Crimes
```{r}
ggplot(merged_data, aes(x = UnemploymentRate, y = TotalCrimes)) +
  geom_point(color = brewer.pal(8, "Set2")[3], size = 3) +  
  geom_smooth(method = "lm", se = FALSE, color = brewer.pal(9, "Blues")[4]) + 
  geom_text(aes(label = Ward), vjust = -1, size = 3, color = "black") +  

  labs(title = "Unemployment Rate vs. Total Crimes by Ward",
       x = "Unemployment Rate (%)",
       y = "Total Crimes",
       caption = "Source: Open Data DC - Crime Incidents Last 30 & Economic Characteristics") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.caption = element_text(hjust = 0.5, size = 10)  # Center the caption
  )


```


Total Crimes in D.C. 
```{r}
total_crimes_dc <- crime_data %>%
  group_by(OFFENSE) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))


ggplot(total_crimes_dc, aes(x = reorder(OFFENSE, Count), y = Count)) +
  geom_col(fill = "steelblue") +  # Set a single color for all bars
  geom_text(aes(label = Count), vjust = -0.3, size = 3.5, color = "black") +
  coord_flip() +
  labs(title = "Total Number of Crimes by Offense Type in Washington, DC",
       x = "Offense Type" ,
       caption = "Source: Open Data DC - Crime Incidents Last 30 Days - https://opendata.dc.gov/datasets/dc3289eab3d2400ea49c154863312434_8/explore?location=38.904042%2C-77.012050%2C10.64") +
  theme_minimal() +
  theme(legend.position = "none", plot.caption = element_text(hjust = 0))

```



Top 3 Crimes by Ward 
```{r}
top_crimes <- crime_data %>%
  group_by(WARD, OFFENSE) %>%
  summarise(Count = n()) %>%
  arrange(WARD, desc(Count)) %>%
  group_by(WARD) %>%
  slice_max(order_by = Count, n = 3) %>%
  mutate(WARD = paste("Ward", WARD))

ggplot(top_crimes, aes(x = Count, y = reorder(OFFENSE, Count), fill = OFFENSE)) +  
  geom_col() +
  facet_wrap(~ WARD, scales = "free_y") +
  scale_fill_viridis_d() +
  labs(title = "Top 3 Crimes by Ward",
       x = "Number of Crimes",  
       y = "Offense Type",
       caption = "Source: Open Data DC - Crime Incidents Last 30 Days - https://opendata.dc.gov/datasets/dc3289eab3d2400ea49c154863312434_8/explore?location=38.904042%2C-77.012050%2C10.64") + 
  theme_minimal() +
  theme(legend.position = "none",
        plot.caption = element_text(hjust = 0)
        )

```


