---
title: "INST_462_Group_Project"
output: html_document
date: "2025-04-15"
---

## Team Members:
Gabriel Gonzalez, Jaida Ries, Joseph Toto, DeJon Young
 

## Roles:
- Gabriel Gonzalez: Developer 
- Jaida Ries: Communicator, Project Manager
- Joseph Toto: Designer
- DeJon Young: Data Engineer

---

## Research Questions — COMMUNICATIONS
*Clear description of research questions, including rationale for why your chosen data should be able to address these questions in some way. Explain who the audience is, and why they might care about what you find*

Our refined research question is:

**"How does income correlate with life expectancy in Washington, DC, by ward?"**

We chose this question because Washington, DC is a city known for stark disparities between neighborhoods. By focusing on income and crime at the ward level, we aim to investigate whether and how economic inequality may be related to safety and well-being.

**Audience**: This analysis is especially relevant for policymakers, public health officials, and local residents who are interested in understanding and addressing disparities in community outcomes across DC.

---
## Data Description — DATA ENGINEER
## Data Links 
*insert links here*
https://opendata.dc.gov/apps/2f6f21d035984e9c9438ae494022aa2a/explore 
https://opendata.dc.gov/datasets/dc3289eab3d2400ea49c154863312434_8/explore?location=38.904042%2C-77.012050%2C10.64 

The **initial datasets** selected for Phase 2 are:

1. **Economic Characteristics of DC Wards.csv**  
   - Contains ward-level economic indicators including median household income, poverty rates, and unemployment levels.
   - Key variables:
     - `Ward`: Unique identifier for each ward.
     - `MedianIncome`: Median household income.
     - `PovertyRate`: Percentage of residents below the poverty line.
     - `UnemploymentRate`: Percentage of the population unemployed.

2. **Crime_Incidents_in_the_Last_30_Days.csv**  
   - Includes crime incidents reported in the last 30 days, categorized by ward.
   - Key variables:
     - `WARD`: Unique identifier for each ward.
     - `OFFENSE`: Type of crime committed.
     - `REPORT_DAT`: Date and time when the crime was reported.

---


## Code -- DEVELOPER
*Your code is in an R markdown file that is clearly organized into sections that are well-explained*
```{r}
library(tidyverse)
library(ggplot2)
```


# Loading and Cleaning Economic Data
```{r}
# Load Economic Data
economic_data <- read_csv("Economic Characteristics of DC Wards.csv") %>%
  select(
    Ward = Name,
    MedianIncome = `INCOME AND BENEFITS (IN 2022 INFLATION-ADJUSTED DOLLARS):  Total households:  Median household income (dollars)`,
    PovertyRate = `PERCENTAGE WHOSE INCOME IN THE PAST 12 MONTHS IS BELOW THE POVERTY LEVEL:  All people`,
    UnemploymentRate = `EMPLOYMENT STATUS:  Civilian labor force:  Unemployment Rate`
  ) %>%
  mutate(Ward = str_extract(Ward, "Ward \\d"))  # Ensure consistent formatting

# Display the first few rows
head(economic_data)
```

# Loading and Aggregating Crime Data
```{r}
crime_data <- read_csv("Crime_Incidents_in_the_Last_30_Days.csv")

# Aggregate Crime Data by Ward 
aggregated_crime <- crime_data %>%
  group_by(WARD) %>%
  summarize(TotalCrimes = n()) %>%
  ungroup()

# Display the aggregated data
head(aggregated_crime)

```

# Code to Modify the Ward Column 
```{r}
economic_data_cleaned <- economic_data %>%
  mutate(Ward = str_extract(Ward, "\\d+"))

# View the cleaned data 
print(economic_data_cleaned)

```

# Merging the Datasets 
```{r}

# Convert Ward in economic_data_cleaned to numeric 
economic_data_cleaned <- economic_data_cleaned %>%
  mutate(Ward = as.numeric(Ward))

# Merge the datasets
merged_data <- economic_data_cleaned %>%
  inner_join(aggregated_crime, by = c("Ward" = "WARD"))

```

```{r}
# View the merged data
print(merged_data)
```


## Initial Plots 
Median Household Income vs. Total Crimes Plot
```{r}
# Plot 1: Median Household Income vs. Total Crimes 
ggplot(merged_data, aes(x = MedianIncome, y = TotalCrimes)) + 
  geom_point(color = "blue", size = 3) + 
  geom_smooth(method = "lm", se = FALSE, color = "darkgray")
  labs(title = "Median Household Income vs. Total Crimes by Ward",
       x = "Median Household Income ($)",
       y = "Total Crimes") + 
  theme_minimal()
```


Poverty Rate vs. Total Crimes Plot
```{r}
#Plot 2: Poverty Rate vs. Total Crimes
ggplot(merged_data, aes(x = PovertyRate, y = TotalCrimes)) +
  geom_point(color = "red", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgray")
  labs(title = "Poverty Rate vs. Total Crimes by Ward",
       x = "Poverty Rate (%)",
       y = "Total Crimes") +
  theme_minimal()
```

Total Crimes by Ward Plot
```{r}
ggplot(merged_data, aes(x = reorder(Ward, TotalCrimes), y = TotalCrimes)) +
  geom_bar(stat = "identity", fill = "green") +
  geom_text(aes(label = TotalCrimes), vjust = -0.5, color = "black") + 
  labs(title = "Total Crimes by Ward",
       x = "Ward",
       y = "Total Crimes") +
  theme_minimal() 

```

Income Distribution by Ward

```{r}
ggplot(merged_data, aes(x = reorder(Ward, MedianIncome), y = MedianIncome)) +
  geom_bar(stat = "identity", fill = "purple") +
  labs(
    title = "Median Income by Ward", 
    x = "Ward",
    y = "Median Income ($)"
  ) +
  theme_minimal() 
```

Unemployment Rate vs. Total Crimes
```{r}

# Plot: Unemployment Rate vs. Total Crimes
ggplot(merged_data, aes(x = UnemploymentRate, y = TotalCrimes)) +
  geom_point(color = "orange", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgray") + # Add trend line
  labs(title = "Unemployment Rate vs. Total Crimes by Ward",
       x = "Unemployment Rate (%)",
       y = "Total Crimes") +
  theme_minimal()

```



Graphs -- GRAPHIC DESIGNER 
*include some initial graphs exploring a couple variables and their relationships*

```{r}

```

By merging **"Economic Characteristics of DC Wards.csv"** and **"Crime_Incidents_in_the_Last_30_Days.csv"**, you can analyze correlations between income, poverty, unemployment, and crime rates at the ward level.
